---
title: "Deprivation in Wales"
output: html_notebook
---

# Deprivation in Wales

The [Welsh Index of Multiple Deprivation](https://gov.wales/welsh-index-multiple-deprivation-full-index-update-ranks-2019) (WIMD) provides data on a number of measures of deprivation in Wales, down to LSOA.

## Compile and clean: import data for each WIMD

Expand the **Data** section to see the [WIMD 2019 Index and domain scores by small area](https://gov.wales/sites/default/files/statistics-and-research/2019-12/wimd-2019-index-and-domain-scores-by-small-area_0.ods) and download that. 

It's an ODS file so we'll use the `readODS` package to import it.

```{r import ods}
wimd19scores <- readODS::read.ods("wimd-2019-index-and-domain-scores-by-small-area_0.ods", sheet = 2)
#For some reason skip = 2 throws an error so we set the colnames to match row 3
colnames(wimd19scores) <- wimd19scores[3,]
#Then remove the unneeded ones
wimd19scores <- wimd19scores[c(-1:-3),]
```

Repeat for the [2014 WIMD](https://gov.wales/welsh-index-multiple-deprivation-full-index-update-ranks-2014) - again downloaded from the data section:

```{r import wimd14}
#This time it's xlsx so we use rio's import function and skip the first couple lines
wimd14scores <- rio::import("150812-wimd-2014-domain-scores-revised-en.xlsx", sheet = 2, skip = 2)
```

And the [2011 WIMD](https://gov.wales/welsh-index-multiple-deprivation-full-index-update-ranks-2011):

```{r import 2011 wimd}
#This time we import directly from the URL and the headers are in the first row
wimd11scores <- rio::import("https://gov.wales/sites/default/files/statistics-and-research/2019-04/wimd-2011-individual-domain-scores-and-overall-index-scores-for-each-lsoa.xls", sheet= 2)
```

Let's check the summaries of those:

```{r summary}
summary(wimd19scores)
summary(wimd14scores)
summary(wimd11scores)
```

## Clean: convert ratings to numeric

The WIMD ratings are treated as strings. They need to be numeric:

```{r scores as numeric}
wimd11scores$`WIMD 2011 score` <- as.numeric(wimd11scores$`WIMD 2011 score`)
wimd14scores$`WIMD 2014 (r)` <- as.numeric(wimd14scores$`WIMD 2014 (r)`)
wimd19scores$`WIMD 2019` <- as.numeric(wimd19scores$`WIMD 2019`)
```



## Combine and context: merge the years

Each dataset has an LSOA code, so we can merge on those. To begin with let's just look at overall scores. We add `all = F` so it only matches where there is a match ([more here](https://www.infoworld.com/article/3454356/how-to-merge-data-in-r-using-r-merge-dplyr-or-datatable.html)):

```{r merge}
#Combine the latest two, matching on LSOA code
wimdcombined <- merge(wimd19scores, wimd14scores, by = "LSOA Code", all = F)
#Add in the 2011 data
wimdcombined <- merge(wimdcombined, wimd11scores, by = "LSOA Code", all = F)
```

We now have 36 columns

```{r colnames combined}
colnames(wimdcombined)
```

Let's simplify:

```{r remove columns}
wimdcompared <- wimdcombined[c(1,2,3,4,16,28)]
```

And calculate some differences:

```{r}
wimdcompared$from14to19 <- wimdcompared$`WIMD 2019` - wimdcompared$`WIMD 2014 (r)`
wimdcompared$from11to14 <- wimdcompared$`WIMD 2014 (r)` - wimdcompared$`WIMD 2011 score`
wimdcompared$from11to19 <- wimdcompared$`WIMD 2019` - wimdcompared$`WIMD 2011 score`
```

```{r}
wimdcompared
```

## Export

Write the results to a csv

```{r write csv}
write.csv(wimdcompared,"wimdcompared.csv")
```

