---
title: "Deprivation in Wales - rankings"
output: html_notebook
---

# Deprivation in Wales

The [Welsh Index of Multiple Deprivation](https://gov.wales/welsh-index-multiple-deprivation-full-index-update-ranks-2019) (WIMD) provides data on a number of measures of deprivation in Wales, down to LSOA.

## Compile and clean: import data for each WIMD

Expand the **Data** section to see the [WIMD 2019 Index and domain ranks by small area](https://gov.wales/sites/default/files/statistics-and-research/2019-12/welsh-index-multiple-deprivation-2019-index-and-domain-ranks-by-small-area.ods) and download that. 

It's an ODS file so we'll use the `readODS` package to import it.

```{r import ods}
wimd19ranks <- readODS::read.ods("welsh-index-multiple-deprivation-2019-index-and-domain-ranks-by-small-area.ods", sheet = 2)
#For some reason skip = 2 throws an error so we set the colnames to match row 3
colnames(wimd19ranks) <- wimd19ranks[3,]
#Then remove the unneeded ones
wimd19ranks <- wimd19ranks[c(-1:-3),]
```

Repeat for the [2014 WIMD](https://gov.wales/welsh-index-multiple-deprivation-full-index-update-ranks-2014) - again downloaded from the data section:

```{r import wimd14}
#This time it's xlsx so we use rio's import function and skip the first couple lines
wimd14ranks <- rio::import("150812-wimd-2014-overall-domain-ranks-each-lsoa-revised-en.xlsx", sheet = 2, skip = 2)
```

And the [2011 WIMD, which has to be exported from StatsWales](https://statswales.gov.wales/Catalogue/Community-Safety-and-Social-Inclusion/Welsh-Index-of-Multiple-Deprivation/Archive/WIMD-2011/wimd2011)

```{r import 2011 wimd}
wimd11ranks <- rio::import("wimd11export.csv", skip = 2)
colnames(wimd11ranks) <- wimd11ranks[1,]
wimd11ranks <- wimd11ranks[c(-1,-2),]
colnames(wimd11ranks)[1] <- "LA"
colnames(wimd11ranks)[2] <- "LSOAandName"
```

Let's check the summaries of those:

```{r summary}
summary(wimd19ranks)
summary(wimd14ranks)
summary(wimd11ranks)
```

## Clean: convert rankings to numeric

The WIMD rankings in two are treated as strings. They need to be numeric:

```{r scores as numeric}
wimd11ranks$`WIMD (1)` <- as.numeric(wimd11ranks$`WIMD (1)`)
wimd19ranks$`WIMD 2019` <- as.numeric(wimd19ranks$`WIMD 2019`)
```

## Clean: extract LSOA codes

The WIMD11 column mixes code and name:

```{r}
head(wimd11ranks$LSOAandName)
```

We need `tidyr` to use the `separate` function to split those out:

```{r}
library(dplyr)
library(tidyr)
```

And specify what we separate it into:

```{r}
wimd11ranks <- wimd11ranks %>% 
  tidyr::separate(LSOAandName, sep=":", into = c("LSOA Code","LSOAname"))
```




## Combine and context: merge the years

Each dataset has an LSOA code, so we can merge on those. To begin with let's just look at overall scores. We add `all = F` so it only matches where there is a match ([more here](https://www.infoworld.com/article/3454356/how-to-merge-data-in-r-using-r-merge-dplyr-or-datatable.html)):

```{r merge}
#Combine the latest two, matching on LSOA code
wimdrankcombined <- merge(wimd19ranks, wimd14ranks, by = "LSOA Code", all = F)
#Add in the 2011 data
wimdrankcombined <- merge(wimdrankcombined, wimd11ranks, by = "LSOA Code", all = F)
```

We now have 46 columns

```{r colnames combined}
colnames(wimdrankcombined)
```

Let's simplify:

```{r remove columns}
wimdrankcombined <- wimdrankcombined[c(1,2,3,4,18,38)]
colnames(wimdrankcombined)
```

Let's clarify the 2011 column:

```{r}
colnames(wimdrankcombined)[6] <- "WIMD 2011 rank"
colnames(wimdrankcombined)
```

And calculate some differences:

```{r}
wimdrankcombined$from14to19 <- wimdrankcombined$`WIMD 2019` - wimdrankcombined$`WIMD 2014 (r)`
wimdrankcombined$from11to14 <- wimdrankcombined$`WIMD 2014 (r)` - wimdrankcombined$`WIMD 2011 rank`
wimdrankcombined$from11to19 <- wimdrankcombined$`WIMD 2019` - wimdrankcombined$`WIMD 2011 rank`
```

```{r}
wimdrankcombined
```

## Export

Write the results to a csv

```{r write csv}
write.csv(wimdrankcombined,"wimdrankcombined.csv")
```

